---
title: "HW3-drafting-viz"
format: 
  html:
    embed-resources: true
editor_options: 
  chunk_output_type: console
---

### Written Responses

#### 1: 
Which option do you plan to pursue?

I still plan to pursue option 1, the infographic.

#### 2:
Restate your questions. Has this changed at all since HW #1? If yes, how so?

My main questions as I described them in homework 1 was if metal content in mammal tissue correlates with proximity to mining areas. If so, what can this tell us about morphological differences in humans, if anything? As I continue to think about this project, my question does generally stay the same, but it also has expanded. I see the flow of my infographic answering these series of questions: What counties have the highest amounts of various metal contents in mammal tissue? What counties have the most proximity to mining sites? Do these things correlate?

#### 3:
Explain which variables from your data set(s) you will use to answer your question(s), and how.

My data set has 40 variables, a lot of which will not be used in my analysis. The most important variables for me are the latitude, longitude, and county for spatial analysis, total_length to scale the metal content by the size of the animal, and the metal of choice. I've used mg so far for my analysis, but I am planning to use fe which might give me more helpful results as it relates to iron ore mining. I've combined this original dataset to another dataset that has all of the mining sites in Minnesota and the county where they are along with the type of mining. I am joining those two by county to plot them all together with the lat and long from the original dataset.

#### 4:
Find at least two data viz that you could (potentially) borrow / adapt pieces from.

#### 5:
Hand drawn visualizations

#### 6:
Mock up visualizations

#### 7:
a. What challenges did you encounter or anticipate encountering as you continue to build / iterate on your visualizations in R? If you struggled with mocking up any of your three visualizations (from #6, above), describe those challenges here.

The main challenges that I faced while creating these visualizations was figuring out how to combine the data when the mining sites data did not have any latitude and longitude data. However, I was able to combine on county. Also,

#### b.
What ggplot extension tools / packages do you need to use to build your visualizations? Are there any that we haven’t covered in class that you’ll be learning how to use for your visualizations?

One important package for my spatial visualization is the tigris package, which allows me to underlay a shapefile of the state of Minnesota.

#### c.
What feedback do you need from the instructional team and / or your peers to ensure that your intended message is clear?

### Code Set up

```{r}
#| eval: true
#| echo: true
# Load libraries
library(dplyr)
library(janitor)
library(ggplot2)
library(here)
library(tmap)
library(sf)
library(hrbrthemes)
library(tigris)
```

```{r}
#| eval: true
#| echo: true
# Read in and clean data
df <- read.csv('data/mammal_hvy_metals_by_pop_withGUIDs.csv') %>%
  clean_names() %>%
  filter(!is.na(latitude) & !is.na(longitude))
# add in more data cleaning post mortem, once I understand what variables I am interested in

mine_locations <- read.csv('data/mine_locations.csv') %>% 
  clean_names()
```

This dataset makes use of the University of Minnesota's Bell Museum of Natural History collection examining specimens of four mammal species (a mouse, shrew, bat and squirrel) to ask how tissue metal content has changed over a 94-year time period (1911-2005), and implications for measures of individual performance (body size and cranial capacity). The metal content of organisms is often elevated closer to cities, so these specimens were examined for spatial variation in metal exposure based on their proximity to human populations and the size of those populated areas at the time of collection. Analysis of mammal tissues focused on six heavy metals associated with human activity (Pb, Cd, Zn, Cu, Cr, Ni, Mn), to address whether these anthropogenic metal pollutants vary in concert with human activity.

```{r}
#| eval: false
#| echo: false
unique(df)
```

### Visualizations

```{r}
#| eval: true
#| echo: true
# First data viz -> collections over time
df %>%
  mutate(mg_prop = mg/total_length) %>%
  ggplot(aes(x = year_collected, y = mg_prop)) +
  geom_area(fill = "#e0e1dd") +
  geom_line(color = "#bde0fe") +
  theme_minimal() +
  labs(
    title = "Magnesium levels in mammal tissue over time",
    subtitle = "Mg level (ppm) is proportioned to animal size using length (mm)",
    x = "Year",
    y = "Mg Level",
    caption = "Data Source: Environmental Data Initiative (EDI), published May 28, 2024. \nCreators come from the University of Minnesota and The Minneapolis-St. Paul Long Term Ecological Research") +
  theme_ipsum() +
  theme(axis.title.x = element_blank())
# the average content of Mg in the body of most animals is about 0.4 Mg / kg of body weight
```

```{r}
#| eval: true
#| echo: true
# Second data viz -> use sf object to plot lat and long --> no longer doing this

# merge 
mines_and_metals <- merge(x=df, y=mine_locations, by = "county", all.y=TRUE) %>% 
  filter(!is.na(fe))

mining_locations <- mines_and_metals

mn_shapefile <- states(cb=TRUE) %>% 
  filter(STUSPS == "MN")

ggplot() +
  geom_sf(data = mn_shapefile, fill = "lightgrey", color = "black") +
  geom_point(data = df, aes(x = longitude, y = latitude)
             # find concerning metals amount to be able to make a gradient
             ) +
  geom_point(data = mining_locations, aes(x = longitude, y = latitude, color = commodity),
             shape = 17, size = 3) +  # Overlay red triangles for mining sites
  labs(
    title = "Mining Locations and Concerning Metal Content in Mammals",
    subtitle = "Concerning metal content is defined as roughly 20 mg/kg",
    caption = "Data Source: Environmental Data Initiative (EDI), published May 28, 2024\nUnited States Geological Survey (USGS) Minerals Yearbook"
    ) +
  guides(fill = guide_legend(title = "Mining type: ")) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid = element_blank(),
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    legend.title = element_blank()
    )

# establish bounding box of msp 
msp_lat_min <- 44.85
msp_lat_max <- 45.15
msp_long_min <- -93.45
msp_long_max <- -93.0

# map bounded to msp area  
ggplot() +
  geom_sf(data = mn_shapefile, fill = "lightgrey", color = "black") +
  geom_point(data = df, aes(x = longitude, y = latitude)
             # find concerning metals amount to be able to make a gradient
             ) +
  geom_point(data = mining_locations, aes(x = longitude, y = latitude, color = commodity),
             shape = 17, size = 3) +  # Overlay red triangles for mining sites
  xlim(msp_long_min, msp_long_max) +
  ylim(msp_lat_min, msp_lat_max) +
  labs(
    title = "Mining Locations and Concerning Metal Content in Mammals",
    subtitle = "Concerning metal content is defined as roughly 20 mg/kg",
    caption = "Data Source: Environmental Data Initiative (EDI), published May 28, 2024\nUnited States Geological Survey (USGS) Minerals Yearbook"
    ) +
  guides(fill = guide_legend(title = "Mining type: ")) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid = element_blank(),
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    legend.title = element_blank()
    )

```

```{r}
#| eval: true
#| echo: true
# Third data viz -> sum magnesium levels by county
mg_df <- df %>%
  group_by(county) %>%
  mutate(total_mg = sum(mg)) %>%
  arrange(total_mg) %>%
  ungroup() %>%
  mutate(county = factor(county, levels = unique(county)))

ggplot(mg_df, aes(x = county, y = total_mg)) +
  geom_point() +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Magnesium levels (in mammal tissue) by county",
    y = "Total Mg"
  )
```

```{r}
#| eval: true
#| echo: true
# Third data viz -> sum magnesium levels by county
mg_df <- df %>%
  group_by(county) %>%
  mutate(total_mg = sum(mg)) %>%
  mutate(mg_prop = mg/total_length) %>%
  ungroup()

ggplot(mg_df, aes(x = county, y = mg_prop)) +
  geom_point() +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Magnesium levels in mammal tissue proportional to animal length",
    y = "Mg level / total length"
  )
```
